---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}

wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")



```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

There is a wide range of pay grades for advisors, even for people who have the same title (i.e. "advisor" or "senior advisor"). Some senior advisors are not paid, and only a handful of advisors make the top salary of $197,200 - which happen to be part-time detailees.The issue with some advisors not getting paid will complicate the analysis a bit. You really can't either exclude them or include those upaid staffers people, because if you exclude them, then there will be questions of who these people are. Similarly, since the White House decided to include these staffers on the report, now the questions are more so about why the White House has unpaid staffers.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |>
  summarize(mean_salary = mean(SALARY), median_salary = median(SALARY)
  )

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |>
  summarize(mean_salary = mean(SALARY), median_salary = median(SALARY), min_salary = min(SALARY), max_salary = max(SALARY)
  )

```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The fact that the average salary is higher than the median salary suggests that the data is not evenly distributed. It could be because the highest-paid staffers ($190,000 +) are bringing the average up, even though most people in the White House are making much less.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I would like to investigate why the White House is hiring unpaid advisors, and why they are being included on the same list with paid employees and advisors. Are these people getting paid in another form of compensation (gifts, favors, potential say on policies), or maybe from another governmental department? My next step would be to try and either talk to the Director of the White House Budget Office or even some of these unpaid advisors and detailees to find out more about what exactly their role (s) in the administration are.

Another interesting idea could be a deep dive on Alice Johnson, the unpaid "pardon czar", who I believe became famous when Trump pardoned her during his first term. It'd be interesting to hear what Johnson's role is like and if her opinions actually carry any merit or if this is just a ceremonial position. 
