---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset
```{r}
sample5 <- mandms |>
sample_n (5)
```

### Task 4: Calculate the mean and standard deviation of the sample, using either the red or black number of M&Ms
```{r}
mean(sample5$red)
sd(sample5$red)
```


### Task 5: Calculate the mean and standard deviation of the population, using the same measure as in Task 4
```{r}
mean(mandms$red)
sd(mandms$red)

```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

Between the two sets, the standard deviations are close than the means are. The standard deviation in the entire population is at about 16.5, while the standard deviation of the sample of 5 red M&Ms is at about 14.8. There is a much bigger difference between the two sets of means, with the mean of the sample of 5 red M&Ms at about 84.4 and the mean of the population at about 73.


### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

The data shows that the sample mean was higher than the population mean, while the sample standard deviation was lower than the population. This shows that a small random sample may not always be a perfect representation of the entire population. Larger samples may actually better show us estimates that are closer to the population statistics.

The sample of five observations had a mean of 84.4 red M&Ms and a standard deviation of 14.77. The entire dataset had a mean of 72.98 and a standard deviation of 16.46. The sample mean was higher than the population mean, while the sample standard deviation was slightly lower. Different random samples could produce completely different means and standard deviations, so this shows that a small random sample may not perfectly represent the entire population. Larger samples may show estimates that are closer to the population statistics.



When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
